Boosting Higgs Pair Production in the $b\bar{b}b\bar{b}$ Channel at the LHC with Multivariate Techniques

### 2015-07-14 paper/sec-mva.tex
work on paper

For the boosted category, we see that substructure variables
are the most helpful ones, in particular the $k_t$-splitting scale
$\sqrt{d_{12}}$ and the subjettiness ratio $\tau_{12}$.
%
The ratio of ECFs $D_2^{(\beta)}$ and
$C_2^{(\beta)}$ also seems to provide
useful discrimination power.
%
Concerning other kinematic variables, the angular
separation between the two large-$R$ jets $\Delta R_{hh}$,
the invariant mass of the di-Higgs system $m_{hh}$ and the various
$p_T$ also have discrimination power.
%
Other variables provide limited information, in particular the invariant
mass of the Higgs candidates, $m_{h_1}$ and $m_{h_2}$: this
is due to the realistic
simulation of detector resolution introduced with the smearing.
%
We have verified that the post-MVA
signal significance in the boosted category is substantially
degraded if substructure variables are not used.

@@

The results of Fig.~\ref{fig:sb_mva} are interesting in that they show
that for the resolved and boosted categories the improvement in signal
significance arising from the MVA is rather moderate, less that a factor two
in the two cases.
%
For no value of the cut the $S/\sqrt{B}$ value in these two categories
goes above one.
%
The real usefulness of the MVA can be seen for the resolved category,
where the signal significance rises from around 1.0 in the cut-based
analysis to above 3.0 for values of the ANN output cut above 0.85.
%
This value of the cut seems reasonable, since we know from Fig.~\ref{fig:nev2}
that the number of signal events at the HL-LHC is still quite substantial,
of the order of 100.
%
Therefore, thanks to the MVA, we predict that one should be able
to find evidence for the production of Higgs boson pairs
at the HL-LHC using the $4b$ channel alone.

Another very important benefit of the MVA applied to the boosted
category is that fact that the background is substantially
reduced so that $S/B$ can be as large as 10\% for the optimal
values of the ANN output cut.
%
Note that this is not true for the other two categories,
where $S/B$ is roughly constant.
%
A reasonably large value of $S/B$ is extremely beneficial
from the experimental point of view, since it indicates
that the measurement is feasible even if the
systematic uncertainties are not very small.

Our final results are summarized in Table~\ref{table:cutflowMVA}, where we indicate
the value of the optimal cut in each category,
the number of signal and background events $N_{\rm ev}$,
and also $S/\sqrt{B}$ and $S/B$.
%
For completeness, we also include the results after the cut-based
analysis.
%
Given that the signal significance is much larger in the boosted as
compared to the other two categories, no additional information
could be obtained by combining the significance of the three
categories, and it seems more efficient to perform
the analysis exclusively in the boosted category.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}[t]
  \centering
  \begin{tabular}{c|c|c|c||c|c|c|c}
    \hline
    \multicolumn{8}{c}{Boosted category}\\
    \hline
    \hline
    \multicolumn{4}{c||}{no MVA cut} & \multicolumn{4}{c}{optimal MVA cut}\\
    \multicolumn{4}{c||}{$y_{\rm cut}=0$} & \multicolumn{4}{c}{$y_{\rm cut}=0.86$}\\
    \hline
    \multicolumn{2}{c|}{$N_{\rm ev}$} &  $S/\sqrt{B}$  & $S/B$
    & \multicolumn{2}{c|}{$N_{\rm ev}$} &  $S/\sqrt{B}$  & $S/B$\\
        Signal & Back   &     &   &  Signal & Back   &     &    \\
    \hline
      193   &  $3.8\,10^4$    &  0.99    &  $5.1\,10^{-3}$  & 107 & 1040 & 3.31  & 0.10\\
        \hline
  \end{tabular}
   $\,$\\
  \vspace{0.4cm}
  \noindent
  \begin{tabular}{c|c|c|c||c|c|c|c}
    \hline
    \multicolumn{8}{c}{Intermediate category}\\
    \hline
    \hline
    \multicolumn{4}{c||}{no MVA cut} & \multicolumn{4}{c}{optimal MVA cut}\\
    \multicolumn{4}{c||}{$y_{\rm cut}=0$} & \multicolumn{4}{c}{$y_{\rm cut}=0.49$}\\
    \hline
    \multicolumn{2}{c|}{$N_{\rm ev}$} &  $S/\sqrt{B}$  & $S/B$
    & \multicolumn{2}{c|}{$N_{\rm ev}$} &  $S/\sqrt{B}$  & $S/B$\\
        Signal & Back   &     &   &  Signal & Back   &     &    \\
    \hline
      660   &   $2.2\,10^6$   &   0.45   &  $3.1\,10^{-4}$ & 480  & $3.2\,10^5$ & 0.85  & $1.5\,10^{-3}$ \\
        \hline
  \end{tabular}
    $\,$\\
  \vspace{0.4cm}
  \noindent
  \begin{tabular}{c|c|c|c||c|c|c|c}
    \hline
    \multicolumn{8}{c}{Resolved category}\\
    \hline
    \hline
    \multicolumn{4}{c||}{no MVA cut} & \multicolumn{4}{c}{optimal MVA cut}\\
    \multicolumn{4}{c||}{$y_{\rm cut}=0$} & \multicolumn{4}{c}{$y_{\rm cut}=0.17$}\\
    \hline
    \multicolumn{2}{c|}{$N_{\rm ev}$} &  $S/\sqrt{B}$  & $S/B$
    & \multicolumn{2}{c|}{$N_{\rm ev}$} &  $S/\sqrt{B}$  & $S/B$\\
        Signal & Back   &     &   &  Signal & Back   &     &    \\
    \hline
    $1.6\,10^{3}$   &   $1.1\,10^{7}$   & 0.50     &  $1.5\,10^{-4}$   & 968 &
    $1.2\,10^{6}$  & 0.90 & $8.2\,10^{-4}$ \\
        \hline
  \end{tabular}
  \caption{\small The results for the number of signal and
    background events
    at the HL-LHC for both the cut-based analysis and after the
    optimal MVA cut, as well as the $S/\sqrt{B}$ and $S/B$
    values in the two cases.
    %
    From top to bottom we show the results for boosted,
    intermediate and resolved categories.
    %
    The optimal value of the MVA cut $y_{\rm cut}$
    is optimised separately in the
    three categories.
    \label{table:cutflowMVA}
  }
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

From Table~\ref{table:cutflowMVA} we see that after the
MVA the signal significance in the boosted category increases
from 0.99 to 3.31, with $S/B$ increasing from $0.005$ to $0.10$,
with still more that 100 signal events at the HL-LHC.
%
Table~\ref{table:cutflowMVA} also shows that for the intermediate
and resolved categories, the improvement due to the use
of the MVA is rather moderate, and signal significance is always
small since backgrounds are still very large.

Table~\ref{fig:sb_mva} contains the main result of this paper:
at the HL-LHC, it should be possible to claim evidence of
Higgs pair production using only the $4b$ final state.
%
Needless to say, to claim discovery and to improve the accuracy
of the measurement of the trilinear coupling it would still
be needed to combine this channel with others like
$b\bar{b}\gamma\gamma$ and $b\bar{b}\tau\tau$, but in any case
it is always important to have solid evidence from
individual channels.

From Table~\ref{fig:sb_mva} it is also possible to
extract the signal significance that would be obtained
at the end of Run II with $\mathcal{L}_{\rm int}=300$ fb$^{-1}$:
in the boosted category we expect 11 signal events and 100 background
events, leading to a $S\sqrt{B}=1.1$, too small specially accounting
for realistic systematic experimental uncertainties.
%
So the full luminosity of the HL-LHC is really needed to exploit
the information contained in this final state.

In the next section we discuss how the expectations for the HL-LHC
summarized in  Table~\ref{fig:sb_mva} are modified in different
conservative and aggressive scenarios for the performance of the
ATLAS and CMS experiments, in particular how the results depend
on the $b$-tagging performance and in the detector momentum
resolution.
%
But before that let's illustrate how the improvement in $S/\sqrt{B}$
obtained from the MVA arises from the power of the jet
substructure variables.

Finally, let us briefly comment on our results would be affected
if we take into account the systematic uncertainty in the measurement
of the backgrounds, in addition to the statistical uncertainty.
%
While there is no unique definition of how signal significance is modified
by experimental systematic uncertainties~\cite{Bityukov:2002eq}, here we estimate
this effect with the simple correction
\be
\frac{S}{\sqrt{B}} \, \to \, \frac{S}{\sqrt{B}} \frac{1}{\sqrt{1+\lp \sigma^{\rm sys}\rp^2\cdot B}} \, ,
\ee
which corresponds to adding in quadrature to the statistical uncertainty  a single
uncorrelated systematic uncertainty $\sigma^{\rm sys}$.
%
Assuming an optimistic value of a total 2\% systematic uncertainty in the
multijet background measurement, we find that the signal significance would be
reduced in the boosted category from 3.31 to 2.8.
%
However, it is beyond the scope of this work to estimate realistically
the systematic uncertainties on QCD multijet distributions (and their bin-by-bin correlation)
that will be achieved at HL-LHC, and in the following we only consider the statistical
uncertainties on the background.
%
But this quick estimate emphasizes the crucial need of an experimentally accurate
measurement of multijet backgrounds at the LHC to fully exploit the $4b$ channel.

>>>>

For the boosted category, we see that substructure variables
are the most helpful ones, in particular the $k_t$-splitting scale
$\sqrt{d_{12}}$ and the subjettiness ratio $\tau_{12}$.
%
The ratios of ECFs $D_2^{(\beta)}$ and
$C_2^{(\beta)}$ also seems to provide
useful discrimination power.
%
Concerning other kinematic variables, the angular
separation between the two large-$R$ jets $\Delta R_{hh}$,
the invariant mass of the di-Higgs system $m_{hh}$ and the various
$p_T$ also have discrimination power.
%
Other variables provide limited information, in particular the invariant
mass of the Higgs candidates, $m_{h_1}$ and $m_{h_2}$: this
is due to the realistic
simulation of detector resolution introduced with the smearing.
%
We have verified that the post-MVA
signal significance in the boosted category is substantially
degraded if substructure variables are not used.

@@

The results of Fig.~\ref{fig:sb_mva} are interesting in that they show
that for the resolved and boosted categories the improvement in signal
significance arising from the MVA is rather moderate, less that a factor two
in the two cases.
%
Even after the MVA, $S/\sqrt{B}$ never grows above 1.0.
%
The real power of the MVA can be seen for the boosted category,
where the signal significance rises from around 1.0 in the cut-based
analysis to above 3.0 for values of the ANN output cut $y_{\rm cut}\ge0.85$.
%
This value of the cut seems reasonable, since we know from Fig.~\ref{fig:nev2}
that the signal yield at the HL-LHC is still quite substantial,
of the order of 100 event.
%
Therefore, thanks to the MVA, we predict that one should be able
to find evidence for the production of Higgs boson pairs
at the HL-LHC using the $4b$ channel alone.

Another important benefit of the MVA applied to the boosted
category is that fact that the background is substantially
reduced so that $S/B$ can be as large as 10\% for the optimal
values of the ANN output cut.
%
Note that this is not true for the other two categories,
where $S/B$ is roughly constant.
%
A reasonably large value of $S/B$ is beneficial
from the experimental point of view, since it indicates
that the measurement is feasible even if the
systematic uncertainties are not very small.

Our final results are summarized in Table~\ref{table:cutflowMVA}, where we indicate
the value of the optimal ANN output
cut in each category,
the number of signal and background events $N_{\rm ev}$ expected
at the HL-LHC,
and also $S/\sqrt{B}$ and $S/B$.
%
For reference, we also include the results at the end of
the cut-based
analysis.
%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}[t]
  \centering
  \begin{tabular}{c|c|c|c||c|c|c|c}
    \hline
    \multicolumn{8}{c}{Boosted category}\\
    \hline
    \hline
    \multicolumn{4}{c||}{no MVA cut} & \multicolumn{4}{c}{optimal MVA cut}\\
    \multicolumn{4}{c||}{$y_{\rm cut}=0$} & \multicolumn{4}{c}{$y_{\rm cut}=0.86$}\\
    \hline
    \multicolumn{2}{c|}{$N_{\rm ev}$} &  $S/\sqrt{B}$  & $S/B$
    & \multicolumn{2}{c|}{$N_{\rm ev}$} &  $S/\sqrt{B}$  & $S/B$\\
        Signal & Back   &     &   &  Signal & Back   &     &    \\
    \hline
      193   &  $3.8\,10^4$    &  0.99    &  $5.1\,10^{-3}$  & 107 & 1040 & 3.31  & 0.10\\
        \hline
  \end{tabular}
   $\,$\\
  \vspace{0.4cm}
  \noindent
  \begin{tabular}{c|c|c|c||c|c|c|c}
    \hline
    \multicolumn{8}{c}{Intermediate category}\\
    \hline
    \hline
    \multicolumn{4}{c||}{no MVA cut} & \multicolumn{4}{c}{optimal MVA cut}\\
    \multicolumn{4}{c||}{$y_{\rm cut}=0$} & \multicolumn{4}{c}{$y_{\rm cut}=0.49$}\\
    \hline
    \multicolumn{2}{c|}{$N_{\rm ev}$} &  $S/\sqrt{B}$  & $S/B$
    & \multicolumn{2}{c|}{$N_{\rm ev}$} &  $S/\sqrt{B}$  & $S/B$\\
        Signal & Back   &     &   &  Signal & Back   &     &    \\
    \hline
      660   &   $2.2\,10^6$   &   0.45   &  $3.1\,10^{-4}$ & 480  & $3.2\,10^5$ & 0.85  & $1.5\,10^{-3}$ \\
        \hline
  \end{tabular}
    $\,$\\
  \vspace{0.4cm}
  \noindent
  \begin{tabular}{c|c|c|c||c|c|c|c}
    \hline
    \multicolumn{8}{c}{Resolved category}\\
    \hline
    \hline
    \multicolumn{4}{c||}{no MVA cut} & \multicolumn{4}{c}{optimal MVA cut}\\
    \multicolumn{4}{c||}{$y_{\rm cut}=0$} & \multicolumn{4}{c}{$y_{\rm cut}=0.17$}\\
    \hline
    \multicolumn{2}{c|}{$N_{\rm ev}$} &  $S/\sqrt{B}$  & $S/B$
    & \multicolumn{2}{c|}{$N_{\rm ev}$} &  $S/\sqrt{B}$  & $S/B$\\
        Signal & Back   &     &   &  Signal & Back   &     &    \\
    \hline
    $1.6\,10^{3}$   &   $1.1\,10^{7}$   & 0.50     &  $1.5\,10^{-4}$   & 968 &
    $1.2\,10^{6}$  & 0.90 & $8.2\,10^{-4}$ \\
        \hline
  \end{tabular}
  \caption{\small The results for the number of signal and
    background events
    at the HL-LHC for both the cut-based analysis and after the
    optimal MVA cut, as well as the $S/\sqrt{B}$ and $S/B$
    values in the two cases.
    %
    From top to bottom we show the results for boosted,
    intermediate and resolved categories.
    %
    The value of the ANN output cut $y_{\rm cut}$
    is optimised separately in the
    three categories.
    \label{table:cutflowMVA}
  }
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

From Table~\ref{table:cutflowMVA} we see that after the
MVA the signal significance in the boosted category increases
from 0.99 to 3.31, with $S/B$ increasing from $0.005$ to $0.10$,
with still more that 100 signal events at the HL-LHC.
%
Table~\ref{table:cutflowMVA} also shows that for the intermediate
and resolved categories, the improvement due to the use
of the MVA is rather moderate, and signal significance is always
small since backgrounds are still  large.
%
The combination of the signal significance of the
three categories leads to a $S/\sqrt{B}=3.53$, and compared
to 3.31 using only the boosted category.

Table~\ref{fig:sb_mva} contains the main result of this paper:
at the HL-LHC, it should be possible to claim evidence of
Higgs pair production using only the $4b$ final state.
%
Needless to say, to claim discovery and to improve the accuracy
of the measurement of the trilinear coupling it would still
be needed to combine this channel with others like
$b\bar{b}\gamma\gamma$ and $b\bar{b}\tau\tau$, but in any case
it is always important to have robust evidence from
individual channels.

From Table~\ref{fig:sb_mva} it is also possible to
extract the signal significance that would be obtained
at the end of Run II with $\mathcal{L}_{\rm int}=300$ fb$^{-1}$:
in the boosted category we expect 11 signal events and 100 background
events, leading to a $S\sqrt{B}=1.1$, too small specially accounting
for realistic systematic experimental uncertainties.
%
So the full luminosity of the HL-LHC is really needed to exploit
the information contained in this final state, unless of course
new BSM dynamics enhance the di-Higgs production rates.

The results of Table~\ref{table:cutflowMVA} include
only the statistical uncertainty in the determination of
the background number of events.
%
We  briefly comment now how these results would be affected
if we also take into account the systematic experimental
uncertainty in the measurement
of the multijet backgrounds distributions.
%
While there is no unique definition of how signal significance is modified
by experimental systematic uncertainties~\cite{Bityukov:2002eq}, here we estimate
this effect with the simple correction
\be
\frac{S}{\sqrt{B}} \, \to \, \frac{S}{\sqrt{B}} \frac{1}{\sqrt{1+\lp \sigma^{\rm sys}\rp^2\cdot B}} \, ,
\ee
which corresponds to adding in quadrature to the statistical uncertainty  a single
uncorrelated systematic uncertainty $\sigma^{\rm sys}$.
%
Assuming an aggressive value of $\sigma^{\rm sys}=2\%$,
we find that the combined signal significance would be
reduced from 3.51 to 3.03.
%
While this quick estimate emphasizes the critical
importance of an experimentally accurate
measurement of multijet backgrounds at the LHC to fully exploit the $4b$
channel,
it is beyond the scope of this work to estimate realistically
the systematic uncertainties on QCD multijet distributions (and their bin-by-bin correlation)
that will be achieved at HL-LHC.

In the next section we discuss how the expectations for the HL-LHC
summarized in  Table~\ref{fig:sb_mva} are modified in different
conservative and aggressive scenarios for the performance of the
ATLAS and CMS experiments, in particular how the results depend
on the $b$-tagging performance and in the detector momentum
resolution.
%
But before that let us illustrate how the improvement in $S/\sqrt{B}$
obtained from the MVA in the boosted
category arises from the information contained in the
jet substructure variables.
